Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]

### 2016-05-26 manuscript/pref_sampling.tex
Complete rough draft of paper; simulations and bearded seal example both worked up

\title{Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage \footnotemark[2]}

@@

\begin{abstract}
 %\noindent {\emph{Summary}}
\small
\begin{enumerate}
\item  Count and occupancy surveys are often used to estimate the density, abundance, and distribution of animal populations. Recently, model-based approaches to analyzing survey data (i.e. species distribution models) have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures. Model-based analysis often makes use of species-covariate relationships and/or spatially autocorrelated random effects to help predict density or occurrence in unsampled locations.

\item Species distribution models often make the implicit assumption that locations chosen for sampling and animal abundance are conditionally independent given modeled covariates.  However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models.  The approach works by jointly modeling animal abundance/occurrence and the locations selected for sampling, and specifying a dependent correlation structure between the two models.

\item Using simulation, we show that under moderate levels of preferential sampling, our modeling approach reduces bias from $\approx$40\% (under a naive species distribution model) to $approx$5\%.

\item Species case study

\item When animal population surveys are conducted without a well defined sampling frame, ecologists should recognize the potential for biasing effects of preferential sampling. Joint models, such as those described in this paper, can be used to test and correct for such biases.
\end{enumerate}

 Word count: XXXX
\end{abstract}

@@

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}

@@

One of the main advantages of using SDMs advanced in the literature is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

Need to acknowledge that "sample selection bias" has been addressed extensively in SDM modeling with presence-only data (e.g. adjusting "background" or "availability" data to be similar to the locations chosen for sampling).

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review preferential sampling bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate utility of our proposed modeling approach by analyzing a data set of MYSTERY SPECIES X.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that preferential sampling is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g., occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.  In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\lambda_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\lambda}=\{ \lambda_1,\lambda_2,\hdots,\lambda_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\lambda_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}
\begin{eqnarray}
  Y_i & \sim & \text{Bernoulli}(Z_i p_i),
\end{eqnarray}
where $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_pois}
  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use standard bracket notation to denote probability mass and density functions.  For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.  We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $Y_i \sim f(\nu_i)$.  We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector taking on a value of $R_i = 1.0$ if sampling unit $i$ is sampled and is zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

\citet{DiggleEtAl2010} demonstrated PS with an environmental monitoring problem, whereby pollutant monitoring stations were more highly clustered around urban areas with high concentrations of pollutants than in rural areas with comparably low levels of pollutants.  Fitting simple geostatistical models without fixed effects led to positively biased estimates of landscape-level pollutant concentrations.  Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g., an ``urbanity" index) would likely reduce or eliminate bias.  However, the primary point of \citet{DiggleEtAl2010} is well taken: inclusion of spatially autocorrelated random effects in a statistical model is insufficient to remove the potentially biasing effects of PS.

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates are only able to explain a portion of variation present in the data.  If the locations selected for sampling are related (intentionally or unintentionally) to some unmodelled factor related to abundance, bias may still occur.  Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One such example is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon,
   \label{eq:Z_cont}
\end{eqnarray}
the relative density of sampling locations at $i$ would be written as
\begin{eqnarray}
   p_i & \propto & \exp(\xi_i + b \mu_i).
   \label{eq:p_cont}
\end{eqnarray}
Here, the parameter $b$ describes the level of preferential sampling; $b=0$ implies no preferential sampling, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq. \ref{eq:Z_cont}, finding that this approach helped reduce bias associated with preferential sampling, but did not perform as well as the full joint model.

@@

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling preferential sampling.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the the spatial process model for density.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, sampling may be more strongly preferential close to bases of operations, (e.g., landing strips in the case of aerial surveys), but less so in areas that are harder to get to.

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then write the link-scale expectation as in eq. \ref{eq:mu}. 
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\eta_i$ are assumed independent of the abundance process.

 The formulation in Eq. \ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the preferential sampling process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.

@@

To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
\begin{eqnarray*}
  N_i \sim \textrm{Poisson}(\mu_i),
\end{eqnarray*}
where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq. \ref{eq:mu}. Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig. \ref{PS_sims}). Specific details of of data generation procedures are presented in Appendix S1.

For each simulated landscape we generated three virtual count surveys using eqs. \ref{eq:R} and \ref{eq:nu}.  Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, not covariate of spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.  In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.  For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.  We refer to these scenarios as moderate and pathological preferential sampling, respectively (see fig. \ref{fig:sim_R}).  Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.

@@

>>>>

\title{Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]}

@@

\begin{abstract}
 %\noindent {\emph{Summary}}
\small
\begin{enumerate}
\item  Wildlife surveys are often used to estimate the density, abundance, or distribution of animal populations. Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.

\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.  However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.  The approach works by jointly modeling wildlife state variables and the locations selected for sampling, and specifying a dependent correlation structure between the two models.

\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce bias.

\item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea. Models that included a preferential sampling effect led to lower estimates of abundance than models without, but were not favored by an integrated Akaike Information Criterion.  Importantly, the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.

\item When wildlife surveys are conducted without a well defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling. Joint models, such as those described in this paper, can be used to test and correct for such biases.  Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

 Word count: 5779
\end{abstract}

@@

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}
\def\bfX{\bf X}
\def\bfbeta{\boldsymbol{\beta}}

@@

One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.  In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  Y_i & \sim & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_pois}
  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use standard bracket notation to denote probability mass and density functions.  For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.  We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $R_i \sim f(\nu_i)$.  We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

\citet{DiggleEtAl2010} demonstrated PS with an environmental monitoring problem, whereby pollutant monitoring stations were more highly clustered around urban areas with high concentrations of pollutants than in rural areas with comparably low levels of pollutants.  Fitting simple geostatistical models without fixed effects led to positively biased estimates of landscape-level pollutant concentrations.  Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g. an ``urbanity" index) would likely reduce or eliminate bias.  However, the primary point of \citet{DiggleEtAl2010} is well taken: inclusion of spatially autocorrelated random effects in a statistical model is insufficient to remove the potentially biasing effects of PS.

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon,
   \label{eq:Z_cont}
\end{eqnarray}
the spatially continuous relative intensity ($\psi_i$) of sampling locations at $i$ could be written as
\begin{eqnarray}
   \psi_i & \propto & \exp(\xi_i + b \mu_i).
   \label{eq:p_cont}
\end{eqnarray}
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq. \ref{eq:Z_cont}, finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

@@

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling PS.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the PS model to be written as a simple linear function of the the spatial process model for density.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, sampling may be more strongly preferential close to bases of operations, (e.g. landing strips in the case of aerial surveys), but less so in areas that are harder to get to.

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then write the link-scale expectation as in eq. \ref{eq:mu}.
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\eta_i$ are assumed independent of the abundance process.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq. \ref{eq:nu} are estimated to permit parameter identification.

 The formulation in Eq. \ref{eq:nu} is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one can contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.

@@

To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
\begin{eqnarray*}
  N_i \sim \textrm{Poisson}(\exp(\mu_i)),
\end{eqnarray*}
where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq. \ref{eq:mu}. Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig. \ref{fig:sim_bias}). Specific details of of data generation procedures are presented in Appendix S1.

For each simulated landscape we generated three virtual count surveys using eqs. \ref{eq:R} and \ref{eq:nu}.  Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.  In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.  For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.  We refer to these scenarios as moderate and pathological PS, respectively (see fig. \ref{fig:sim_R}).  Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.

@@

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{bearded_ests.pdf}
\caption{Aerial survey counts and estimated apparent abundance of bearded seals in the eastern Bering Sea, April 10-16, 2012. Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders the Alaska, USA mainland to the east. In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.  Apparent bearded seal abundance estimates (B) are presented from the model with the lowest integrated AICc score, which included covariate effects but no preferential sampling effect.  Apparent abundance estimates are uncorrected for imperfect detection or species misclassification.}
\label{fig:bearded_ests}
\end{center}
\end{figure*}
